# Logistic Analysis - week 3

*I have learned how to merge datasets from R, apply logistic regression, interpret its output, predict power of the model and conduct cross-validation.*



## Task 2

```{r}
setwd("D:/New folder/OneDrive for Business 1/Git/IODS/IODS-project/data")

alc <- read.table(file = "alc.txt",
                           header = TRUE,
                           dec = ".")
```

### Libraries

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(GGally)
```

### Structure

```{r}
str(alc)
names(alc)
```

This data approach alcohol consumption by students in secondary education of two Portuguese schools. The data attributes include student grades, demographic, social and school related features, and it was collected by using school reports and questionnaires. The response variable (high_use) is binary, and according to hypotheses four covariates were chosen to model its probability: 1-"address" (factor with 2 levels), "freetime" (integer), "sex" (factor with 2 levels) and "failures" (integer).
Source: Paulo Cortez, University of Minho, Guimarães, Portugal, http://www3.dsi.uminho.pt/pcortez 

## Task 3

### Hypotheses

1.[address] The probability of higher alcohol consumption is higher for students living in rural areas;
2.[free time] The probability of higher alcohol consumption is higher for students with more free time;
3.[sex] The probability of higher alcohol consumption is higher for male students;
4.[failures] The probability of higher alcohol consumption is higher for students with more failures.

## Task 4

### Relationships

```{r}
par(mfrow = c(2, 2), mar = c(5, 5, 2, 2), cex.lab = 1.5)
boxplot(alc$alc_use ~ address,
        varwidth = TRUE, 
        data = alc,
        xlab     = "address",
        ylab     = "alcohol use")

boxplot(alc$alc_use ~ freetime,
        varwidth = TRUE, 
        data = alc,
        xlab     = "freetime",
        ylab     = "alcohol use")

boxplot(alc$alc_use ~ sex,
        varwidth = TRUE, 
        data = alc,
        xlab     = "sex",
        ylab     = "alcohol use")

boxplot(alc$alc_use ~ failures,
        varwidth = TRUE, 
        data = alc,
        xlab     = "failures",
        ylab     = "alcohol use")
```

On average students living in rural areas drink more than students living in urban areas. Alcohol use seems to increase with the amount of free time. Male students drink on average more than female students. Students who failed at least once seem more likely to drink more alcohol than those who have never failed, but the amount of failures does not seem to matter for alcohol consumption.

## Task 5

### Fit the model
```{r}
m1 <- glm(high_use ~ address + freetime + failures + sex, data = alc, family = "binomial")
summary(m1)
drop1(m1, test="Chi")
```


Students living in urban areas tend to have lower probability of high alcohol consumption, but as the significance is borderline we drop this covariate. The probability of high alcohol consumption is higher for students with more freetime and failures, as well as for guys compared to girls.

```{r}
m2 <- glm(high_use ~ freetime + failures + sex, data = alc, family = "binomial")
summary(m2)
```

* Probability of high consumption for females 
pi_i =exp(-2.1678 + 0.2509*freetime_i + 0.4696*failures_i) / 1 + exp(-2.1678 + 0.2509*freetime_i + 0.4696*failures_i)


* Probability of high consumption for males 
pi_i =exp(-2.1678 + 0.7333 + 0.2509*freetime_i + 0.4696*failures_i) / 1 + exp(-2.1678 + 0.7333 + 0.2509*freetime_i + 0.4696*failures_i)      


### odds ratios (OR)
```{r}
OR<- coef(m2) %>% exp
```

### confidence intervals (CI)
```{r}
CI<-confint(m2) %>%exp
```

### odds ratios with their confidence intervals
```{r}
cbind(OR, CI)
```
Odds are all higher than 1, meaning that all covariates are positively associated with high alcohol consumption.

## Task 6

### Predictive power of model

```{r}
probabilities <- predict(m2, type = "response")
alc <- mutate(alc, probability = probabilities)
alc <- mutate(alc, prediction=probability>0.5)
```

### tabulate the target variable versus the predictions
```{r}
table(high_use = alc$high_use, prediction = alc$prediction)
table(high_use = alc$high_use, prediction = alc$prediction)%>%prop.table()%>%addmargins()
```
The practice data (actual values) comprised 259 FALSE (low consumption) and 9 TRUE (high consumption), whereas the test (prediction) comprised 100 FALSE and 14 TRUE values.

### loss function (mean prediction error)
```{r}
loss_func <- function(class, prob) {
  n_wrong <- abs(class - prob) > 0.5
  mean(n_wrong)
}
```

### average number of wrong predictions
```{r}
loss_func(class = alc$high_use, prob = alc$probability)
```
28% of individuals are inaccurately classified by the model (m2)

## Bonus

### K-fold cross-validation

```{r}
library(boot)
cv1 <- cv.glm(data = alc, cost = loss_func, glmfit = m1, K = 10)
cv1$delta[1]
```
My model does not provide a smaller prediction error than DataCamp model.

```{r}
m3 <- glm(high_use ~sex*absences, data = alc, family = "binomial")
summary(m3)
cv3 <- cv.glm(data = alc, cost = loss_func, glmfit = m3, K = 10)
cv3$delta[1]
```
A model with an interaction between absences and sex provides smaller prediction error.






