# Clustering and classification - week 4

## Task 1

### Libraries
```{r}
library(MASS)
library(corrplot)
```

### Load the data
```{r}
data("Boston")
```

### Structure and dimension

```{r}
str(Boston)
dim(Boston)
```

### Description
This data frame contains information about housing in the area of Boston.It has 506 rows and 14 columns.

Sources:
https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html

Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. < em >J. Environ. Economics and Management < b >5, 81-102.

Belsley D.A., Kuh, E. and Welsch, R.E. (1980) < em >Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.

## Task 3

```{r}
cor_matrix<-cor(Boston)
corrplot(cor_matrix,type="upper")
```

According to the correlation matrix, the strongest correlations were:

- mean of distances to five Boston employment centres negatively correlates (-0.75) with proportion of owner-occupied units built prior to 1940; with nitrogen oxides concentration (-0.77); with proportion of non-retail business acres per town (-0.71).

- median value of owner-occupied homes negatively correlates (-0.74) with lower status of the population and with average number of rooms per dwelling (0.70).

- full-value property-tax rate positively correlates (0.91) with index of accessibility to radial highways
and with proportion of non-retail business acres per town (0.72).

- proportion of owner-occupied units built prior to 1940 positively correlates (0.73) with nitrogen oxides concentration.

- nitrogen oxides concentration positively correlates with proportion of non-retail business acres per town (0.76)

## Task 4

```{r}
boston_scaled <- scale(Boston)
summary(boston_scaled)
summary(Boston)
```

The means are 0, and the min. and max. (range) are rather smaller in absolute values. 

```{r}
bins <- quantile(boston_scaled[,1])

# create a categorical variable 'crime'
crime <- cut(boston_scaled[,1], breaks = bins, include.lowest = TRUE, label=c("low","med_low","med_high","high"))

# look at the table of the new factor crime
table(crime)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)

# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)

str(train)
str(test)
```


## Task 5 and 6

```{r}
# linear discriminant analysis
lda.fit <- lda(crime~., data = train)

# print the lda.fit object
lda.fit

# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "red", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col=classes,pch=classes)
lda.arrows(lda.fit, myscale = 1.5)


# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
table(correct = correct_classes, predicted = lda.pred$class)

```


## Task 7

```{r}
boston_scaled2 <- scale(Boston)
summary(boston_scaled2)

# euclidean distance matrix
dist_eu <- dist(boston_scaled2,method="euclidean")

# look at the summary of the distances
summary(dist_eu)

# calculate the total within sum of squares
twcss <- sapply(1:10, function(k){kmeans(boston_scaled2, k)$tot.withinss})

```
qplot didn't want to work in Rmarkdown.

**twcss <- sapply(1:10, function(k){kmeans(boston_scaled2, k)$tot.withinss})**

**qplot(x = 1:10, y = twcss, geom = 'line')**

```{r}

# k-means clustering
km <-kmeans(boston_scaled2, centers = 2)

# plot the Boston dataset with clusters
pairs(boston_scaled2, col = km$cluster)
```

According to k-means clustering, the observations of all covariates included in the Boston dataset are somehow clustered in two heterogeneous samples














