# Analysis of longitudinal data - week 6

## Analysis of RATS data
- According to chapter 8 from MABS
```{r}
setwd("D:/New folder/OneDrive for Business 1/Git/IODS/IODS-project/data")
RATSL <- read.table(file = "RATSL.txt",
                           header = TRUE,
                           dec = ".")
RATSL$ID <- factor(RATSL$ID)
RATSL$Group <- factor(RATSL$Group)
str(RATSL)
```



### Libraries

```{r}
library(dplyr)
library(ggplot2)
library(lme4)
```

### Data exploration

```{r}
ggplot(RATSL, aes(x = Time, y = Weight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme_bw() +geom_line(aes(color=ID))+
  scale_y_continuous(limits = c(min(RATSL$Weight), max(RATSL$Weight)))
```


This graph shows the individual weight of subjects by treatment group.
The weight seems to generally increase with time for individuals in every group treatment.
One individual from each group seems to either have higher or lower weight compared to individuals of the same group. From the graph You can notice that their initial weight was already very different  compared to other individuals. This shows that individuals' starting weight should be taken into account when quantifying the treatment effects. There are more individuals in group 1, and they appear to have the lowest weight.  

Lets see if we can visualize the variability better with standardized data.
```{r}
RATSL <- RATSL %>%
  group_by(Time) %>%
  mutate(stdweight = (Weight - mean(Weight))/sd(Weight)) %>%
  ungroup()


ggplot(RATSL, aes(x = Time, y = stdweight, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme_bw() +geom_line(aes(color=ID))+
  scale_y_continuous(limits = c(min(RATSL$stdweight), max(RATSL$stdweight)))
```

Shows what we already knew.

Let's now plot the mean effect of the 3 treatments in time.

```{r}

n <- RATSL$Time %>% unique() %>% length()

# Summary data with mean and standard error of Weight by Group and Time 
df <- RATSL %>%
  group_by(Group, Time) %>%
  summarise( mean = mean(Weight), se = sd(Weight)/sqrt(n)) %>%
  ungroup()


ggplot(df, aes(x = Time, y = mean, linetype = Group, shape = Group)) +
  geom_line() +
  geom_point(size=3) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) + theme_bw() + geom_line(aes(color=Group))+
  scale_y_continuous(name = "mean (Weight) +/- se (Weight)")
```

Here we notice that individuals of Group 3 had the lowest weight throughout the treatment. But then again, they had the lowest weights prior to the treatment. We can also see that the standard error of the mean was higher for the observations made in group 2, probably because of the individual who had a much higher weight.

Time for a quick and dirty analysis. 
Let's average the all repeated observations taken from individual, so we "capture" the general effect of time in each individual.

```{r}
df1 <- RATSL %>%
    group_by(Group, ID) %>%
  summarise( mean=mean(Weight) ) %>%
  ungroup()
```

We can visualize the differences between groups directly.

```{r}
ggplot(df1, aes(x = Group, y = mean))+
geom_boxplot()+
theme_bw() + theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+
stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white")+
scale_y_continuous(name = "mean (Weight)")
```

Based on this graph, it looks like there's an outlier in group 2. I wouldn't remove it as there are too few variables (n=16). I think that the linear mixed model in this scenario would be able to take into account the specific variance between individuals.

Now we fit the linear model

```{r}
fitRATS <- lm(mean ~ Group, data = df1)
anova(fitRATS)
```

From the anova table we see that the group variable (reflecting treatments) significantly affects the weight of the rats.

```{r}
summary(fitRATS)
```

From the summary table we see that treatment 3 (group 3) produces the highest weights, whereas treatment 1 (group 1) produces the lowest weights. 

## Analysis of BPRS data
- According to chapter 9 from MABS

```{r}
setwd("D:/New folder/OneDrive for Business 1/Git/IODS/IODS-project/data")
BPRSL <- read.table(file = "BPRSL.txt",
                    header = TRUE,
                    dec = ".")
BPRSL$treatment <- factor(BPRSL$treatment)
BPRSL$subject <- factor(BPRSL$subject)
str(BPRSL)
```


Let's plot bprs versus the time, ignoring the fact that repeated measurements were done in each individual.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, group = subject))+
geom_text(aes(label = treatment))+
scale_y_continuous(name = "bprs")+
theme_bw()+
theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())
```

It doesn't look like the treatments differ. To be sure, let's run a quick linear model.

```{r}
fitBPRS <- lm(bprs ~ week + treatment, data = BPRSL)
anova(fitBPRS)
```

No evidence was found to say the treatment had an effect on the bprs. However, the week significantly affected bprs.

```{r}
summary(fitBPRS)
```

We can see the coefficients for the model. Treatment 2 had a positive effect on bprs, but non significant, whereas week had a negative significant effect.

```{r}
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme_bw() +geom_line(aes(color=subject))+
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```

From this graph we can see there's a large variability between subjects, making it hard to differentiate treatment effects. But we can indeed see the effect of week.

Let's how the week variables correlate from the wide RATS data.

```{r}
BPRS<-read.table("https://raw.githubusercontent.com/KimmoVehkalahti/MABS/master/Examples/data/BPRS.txt",
                 header=T, sep=" ")
pairs(BPRS[, 3:11], cex = 0.7)
```

The weeks strongly correlate with each other.

Now we fit the linear mixed effect model with individuals as random intercepts, so the model accounts for the variability in subjects.

```{r}
mixedBPRS <- lmer(bprs ~ week + treatment + (1 | subject), data = BPRSL, REML = FALSE)
summary(mixedBPRS)
```

The fixed effects were the same found by the linear model, but the mixed model would allow for the intercept to be correct 20 times (one per individual). Ideally, the data should've been coded in a way that there would be 40 individuals. Now it thinks that each individual received both treatments, probably imposing some correlation between them. but I am no sure.
The standard error is slightly smaller, so we can still try to improve the model by fitting a random intercept and model, so that the slope of each subject can also change according to week.

```{r}
mixedBPRS1 <- lmer(bprs ~ week + treatment + (week| subject), data = BPRSL, REML = FALSE)
summary(mixedBPRS1)
```

Not so much difference. We can compare them by:
```{r}
anova(mixedBPRS,mixedBPRS1)
```

According to the  likelihood ratio test, the random intercept and slope model is better (LRT=7.27, df=2,p=0.03).

At last, we fit a random intercept and slope model that allows for a treatment × week interaction, so the slope  of week will change based on treatment.

```{r}
mixedBPRS2 <- lmer(bprs ~ week *treatment + (week| subject), data = BPRSL, REML = FALSE)
summary(mixedBPRS2)
```

Next we check if the interaction is significant.

```{r}
drop1(mixedBPRS2, test="Chi")
```

The interaction was not significant for predicting pbrs (LRT=3.17,df=1,p=0.07).

Let's see what the model predicts.

```{r}
Fitted <- fitted(mixedBPRS2)
BPRSL <- BPRSL %>% mutate(Fitted)

ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme_bw() +geom_line(aes(color=subject))

ggplot(BPRSL, aes(x = week, y = Fitted, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme_bw() +geom_line(aes(color=subject))
```

In the graph above we can see the negative effect of week on bprs, and that there's not much difference between treatments. The fitted values include random effects, taking into account the variability within individuals across the weeks.