# Analysis - week 2

*I have learned how to organize datasets from R, visualize data, check relationships between variables, apply linear regression, interpret its output and conduct model validation using graphs*

## Tools: learning2014 dataset
```{r}
setwd("D:/New folder/OneDrive for Business 1/Git/IODS/IODS-project/data")

learning2014 <- read.table(file = "learning2014.txt",
                           header = TRUE,
                          dec = ".")
```

## Data Exploration 
```{r}
dim(learning2014)
str(learning2014)
```

The relationship between learning approaches and students achievements in an introductory statistics course in Finland. Learning approaches were attitude, deep, strategic, and surface knowledge. Achiviements measured in points. Surveys were divided by gender (M, F), and age was recorded as well.

Libraries
```{r}
library(GGally)
library(ggplot2)
```

Relationships

- Gender is unbalaned, more males were sampled. 
- The age of male respondents has wider range.
- Possible linear relationship between points and attitude.
- Possible exponential relatonship between points and stra.
- Points seem to vary with gender.
- Maybe attitude and stra are collinear with gender.


```{r}
p <- ggpairs(learning2014, mapping = aes(col=gender, alpha=0.3), lower = list(combo = wrap("facethist", bins = 20)))
p
```


## Fitting the model
```{r}
my_model <- lm(points ~  attitude + stra + surf, data = learning2014)
summary(my_model)
```

According to the model, attitude has a positive effect on points, stra has a non significant positive effect, whereas surf has a non significant negative effect on points. Therefore, stra and surf do not seem to contribute to the prediction of points and can be removed from the model.

```{r}
my_model2 <- lm(points ~  attitude , data = learning2014)
summary(my_model2)
```

Points = 11.6 + 3.5*attitude + noise.
Points are increased on average 3.5 units per unit of change in attitude.
In a scale of 0-100%, attitude explains about 18.6% of the variance in points. It's very little.

## Model validation
```{r}
par(mfrow = c(2,2))
plot(my_model2, which =c(1,2,5))
```

- Normality of the errors - There's deviation from the line, making errors not normally distributed.
- Constant varaince of errors - size of residuals decreases towards the end when fitted values increase.
- Leverage - A couple of observations stand out from the rest (71, 35).



